---
title: "Kinship Ordination"
author: "Justin Van Ee"
date: "2024-02-09"
output: pdf_document
---

# Set up 

```{r, setup}

###
### Packages
###

library(abind)
library(bayesplot)
library(BMTME)
library(Boom)
library(cmdstanr)
library(coda)
library(furrr)
library(doParallel)
library(ggrepel)
library(ggh4x)
library(factor.switching)
library(Matrix)
library(mcmcse)
library(patchwork)
library(posterior)
library(qtl2)
library(rsample)
library(Rcpp)
library(RcppArmadillo)
library(RcppCNPy)
library(readxl)
library(scatterpie)
library(tidyverse)
library(wesanderson)

### Set Directory (Adjust as needed)
dir <- "/Users/justinvanee/Library/Mobile Documents/com~apple~CloudDocs/Documents/bromecast/LVM/Kinship Ordination"

### Load MCMC algorithms from Rcpp
source(file.path(dir, "algorithms", "functions.R"))
```

# Toy Example 

```{r, toy example for method demonstration}

###
### Simulate Data
###

data <- 
  simulate_data(
    G = 20, # genotypes 
    T = 5,  # traits 
    E = 5,  # environments   
    P = 1,  # covariates  
    percent_remove = 0.1, # percent of GxE combinations to remove for out-of-sample prediction 
    trait_cor = 0.95,     # off-diagonal elements of trait correlation matrix  
    dir = dir
  )

###
### Fit and Compare Models 
###

### See functions for details 
# KO: Kinship ordination 
# MMM: Multivariate mixed model 
# UO: Unconstrained ordination 
# CO: Constrained ordination
# CCO: Concurrent ordination
metrics_tib <- 
  simulation_study(
    data=data,
    models = c("KO", "MMM", "UO", "CO", "CCO")               
)

# --- Add Effective Sampling Rate (ESR) ---
metrics_tib <- metrics_tib %>%
  mutate(
    esr = ess / time,
    esr_b = ess_b / time,
    esr_Y = ess_Y / time
  )

# --- Pivot longer for plotting ---
metrics_long <- metrics_tib %>%
  pivot_longer(
    cols = c(
      coverage, bias, sd, cor, esr, 
      coverage_b, bias_b, sd_b, cor_b, esr_b, 
      coverage_Y, bias_Y, sd_Y, cor_Y, esr_Y),
    names_to = "metric",
    values_to = "value"
  )  %>%
  mutate(
    goal = case_when(
      str_detect(metric, "_Y$")  ~ "Posterior Prediction",
      str_detect(metric, "_b$")  ~ "Beta Estimation",
      TRUE                        ~ "Heritability Estimation"
    ),
    # remove the suffixes for clean metric names
    metric = str_remove(metric, "_Y$|_b$")
  )

# --- Relabel and reorder metrics and models ---
metrics_long <- metrics_long %>%
  mutate(
    metric = recode(metric,
      coverage = "Coverage",
      bias = "Bias",
      sd = "Variability",
      cor = "Correlation",
      esr = "Efficiency",
      rmse = "RMSE"
    ),
    metric = factor(metric, levels = c("Efficiency", "Coverage", "Bias", "Variability", "Correlation", "RMSE")),
    goal = factor(goal, levels = c("Heritability Estimation", "Posterior Prediction", "Beta Estimation")),
    model = factor(model, levels = c("KO", "MMM", "UO", "CO", "CCO"))
  )

# --- Define color palette ---
wes_colors <- wes_palette("Darjeeling1", 5, type = "discrete")

# --- Heritability Estimation ---
p1 <- metrics_long %>%
  filter(goal == "Heritability Estimation", 
         metric != "RMSE") %>%
  ggplot(aes(x = model, y = value, color = model, group = metric)) +
  geom_point(size = 3) +
  geom_line(aes(group = model), linewidth = 1, alpha = 0.6) +
  facet_wrap(~ metric, scales = "free_y", nrow = 1) +
  scale_color_manual(values = wes_colors) +
  theme_bw(base_size = 14) +
  labs(
    x = "Model",
    y = "Value",
    color = "Model",
    title = "Heritability Estimation"
  ) +
  theme(
    strip.background = element_rect(fill = "grey95", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

# --- Beta Estimation ---
p2 <- metrics_long %>%
  filter(goal == "Beta Estimation", 
         metric != "RMSE") %>%
  ggplot(aes(x = model, y = value, color = model, group = metric)) +
  geom_point(size = 3) +
  geom_line(aes(group = model), linewidth = 1, alpha = 0.6) +
  facet_wrap(~ metric, scales = "free_y", nrow = 1) +
  scale_color_manual(values = wes_colors) +
  theme_bw(base_size = 14) +
  labs(
    x = "Model",
    y = "Value",
    color = "Model",
    title = "Beta Estimation"
  ) +
  theme(
    strip.background = element_rect(fill = "grey95", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

# --- Posterior Prediction ---
p3 <- metrics_long %>%
  filter(goal == "Posterior Prediction") %>%
  ggplot(aes(x = model, y = value, color = model, group = metric)) +
  geom_point(size = 3) +
  geom_line(aes(group = model), linewidth = 1, alpha = 0.6) +
  facet_wrap(~ metric, scales = "free_y", nrow = 1) +
  scale_color_manual(values = wes_colors) +
  theme_bw(base_size = 14) +
  labs(
    x = "Model",
    y = "Value",
    color = "Model",
    title = "Posterior Prediction"
  ) +
  theme(
    strip.background = element_rect(fill = "grey95", color = NA),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

# Display separately
p1 / p2 / p3 # stacked 

```



# Heatmap of kinship 

```{r, heatmap of kinship}


# Load the example DO mouse data from the qtl2 package
dirpath <- system.file("extdata", "iron.zip", package = "qtl2")
cross <- read_cross2(dirpath)

# Calculate genotype probabilities with an error probability of 0.002
probs <- calc_genoprob(cross, map = cross$gmap, error_prob = 0.002)

# Calculate the kinship matrix using the overall method
kinship <- calc_kinship(probs)

# Define a custom color palette (white → red)
my_colors <- colorRampPalette(c("white", "red"))(100)

# Visualize the kinship matrix for the first 50 individuals
heatmap(cov2cor(kinship[1:7, 1:7]),
        symm = TRUE,
        labRow = NA,    # removes row labels
        labCol = NA,    # removes column labels
        col = my_colors)
```



# Prior Predictive Checks Heritability 

```{r, Prior Predictive Checks Heritability }

# Values of q to explore
q_values <- c(1, 5, 10, 20)

# Simulate for each q and prior type
prior_data <- expand.grid(q = q_values,
                          hierarchical = c(TRUE, FALSE)) %>%
  mutate(h2 = map2(q, hierarchical, ~ simulate_h2_prior(q = .x, hierarchical = .y))) %>%
  unnest(h2) %>%
  mutate(
    prior_type = ifelse(hierarchical, "Hierarchical", "Independent"),
    prior_type = factor(prior_type, levels = c("Independent", "Hierarchical"))
  )

# Facet grid plot
heritability_prior <-
ggplot(prior_data, aes(x = h2), color = "black") +
  geom_histogram(bins = 50, position = "identity", alpha = 0.6, color = "black") +
  facet_grid(prior_type ~ q) +
  theme_bw(base_size = 14) +
  labs(
    x = expression(h^2),
    y = "Count",
    fill = "Prior Type"
  )

# Save plot 
ggsave(plot=heritability_prior, 
       filename=file.path(dir, "Figures", "heritability_prior.png"),  
       dpi=1000, height=5, width=10)

```


# Data Wrangling (Cheatgrass: Growth Chamber + Common Gardens + Physiology)

```{r, Real Data Preparation (Cheatgrass: Growth Chamber + Common Gardens + Physiology)} 

### 
### Data (Growth Chamber)
###

growth_chamber_all <- 
  read.csv(file.path(dir, "Data", "BRTE.data.csv")) %>%
  ### Survival/Survival trait of 2-3 replicates in a different file
  left_join(
    readxl::read_xlsx(file.path(dir, "Data", "CheatgrassSeeds_psu2021growout_data.xlsx")),
    by=c("PopNum", "NewSiteCode")
  ) %>%
  mutate(
    N_survived=as.numeric(N_survived),
    Survival=N_survived/N_planted,
    Survival=if_else(is.na(Survival) & !is.na(N_planted), 0, Survival) ### These six genotypes didn't even germinate and therefore didn't survive 
  )

Y_growth_chamber <- 
  growth_chamber_all %>%
  dplyr::select("FT"="DTF", 
                "Seed(n)"="seedNumber",
                "Seed(m)"="IndividualSeedMass",
                "Biomass"="biomass",
                "Tiller(n)"="tillersN",
                "Height"="height",
                "Growth"="growth",
                "Leaves(n)"="leavesN", 
                "Inflorescence(h)"="panicleHeight",
                "Survival") %>% ### set traits names equal to what we observe in Figure 4 of Gamba manuscript
  as.matrix() #%>%
  #scale()

### Get rows with at least one trait not NA 
keep_idx=rowSums(is.na(Y_growth_chamber))<ncol(Y_growth_chamber)
growth_chamber=growth_chamber_all[keep_idx,]
Y_growth_chamber=Y_growth_chamber[keep_idx,]

### Genotypes Growth Chamber 
genotypes_growth_chamber <- growth_chamber$NewSiteCode

###
### Data (Common Gardens)
###

# Read in data that matches kinship matrix position and genotype ID
kinshipIDs <- read.csv(file=file.path(dir, "Data","93cg_genotypes.csv"))

### This has all traits for all 4 common garden sites from 2022
allsites_flower_harvest <-
  read.csv(file=file.path(dir, "Data","cg_cleandata.csv")) %>%
# Subset trait data for only genotypes that we have kinship data for
  filter(genotype %in% kinshipIDs$genotype) # Right now this only drops 329 plants total and 4 genotypes

### Genotypes Common Gardens  
genotypes_common_gardens <- 
  kinshipIDs %>%
  filter(genotype %in% unique(allsites_flower_harvest$genotype)) %>%
  pluck("NewSiteCode") %>%
  sort()
  
###
### Collect genotypes
###

genotypes_all <-
  tibble(NewSiteCode=unique(c(genotypes_growth_chamber, genotypes_common_gardens))) %>%
  left_join(growth_chamber_all, by="NewSiteCode") %>%
  arrange(NewSiteCode) %>%
  filter(!is.na(id)) # 3 genotypes not sequenced (i.e., no kinship data)
dim(genotypes_all)

### Extract dominant ancestry of each genotype 
Ancestry_labels=c("Mediterranean/Middle East", "Northern Europe", "Eastern Europe", "Central Europe")
Ancestry <- genotypes_all %>%
    mutate(across(V1:V4, round, digits=2),
           V_max=purrr::pmap_dbl(list(V1, V2, V3, V4), max)) %>%
    mutate(Ancestry=case_when(V1==V_max~Ancestry_labels[1],
                              V2==V_max~Ancestry_labels[2],
                              V3==V_max~Ancestry_labels[3],
                              V4==V_max~Ancestry_labels[4])) %>%
    dplyr::select(Ancestry, "NewSiteCode", "V1":"V4") %>%
    mutate(Ancestry_num=as.numeric(as.factor(Ancestry)))


# Read in data that matches kinship matrix position and genotype ID
# Read in kinship matrix for 307 genotypes 
kinship <- read.table(file.path(dir, "Data","BRTE307_IBSmatrix.txt"), sep=",")
K_joint <- as.matrix(kinship[genotypes_all$id,genotypes_all$id])

# Put genotype numbers on rows and columns
rownames(K_joint) <- as.factor(genotypes_all$NewSiteCode)
colnames(K_joint) <- as.factor(genotypes_all$NewSiteCode)

###
### Back to growth chamber data manipulations 
###

Y_growth_chamber <- 
  growth_chamber_all %>%
  filter(NewSiteCode %in% genotypes_all$NewSiteCode) %>% ### This ensures we add the extra 2 genotypes observed in common gardens but not in grennhouse 
  arrange(NewSiteCode) %>%
  dplyr::select("FT"="DTF", 
                "Seed(n)"="seedNumber",
                "Seed(m)"="IndividualSeedMass",
                "Biomass"="biomass",
                "Tiller(n)"="tillersN",
                "Survival",
                "Height"="height",
                "Growth"="growth",
                "Leaves(n)"="leavesN", 
                "Inflorescence(h)"="panicleHeight") %>% ### set traits names equal to what we observe in Figure 4 of Gamba manuscript
  mutate(`Seed(n)`=log(`Seed(n)`),
          Biomass=log(Biomass),
         `Tiller(n)`=log(`Tiller(n)`)) %>% ### Log transform right skewed distribution related to fitness 
  as.matrix() #%>% # Doing scaling for Y_growth_chamber and Y_common_gardens combined 
  #scale()

###
### Back to Common gardens data manipulations 
###

### Calculate Seed(m) and Biomass and set all attributes to NA for plants that didn't survive to Survival
allsites_flowered <- 
  allsites_flower_harvest %>%
  filter(year == 2022) %>%
  mutate(Survival=if_else(seed_count_total>0, 1, 0),
         "FT"=if_else(Survival==0, NA, first_flower),
         "Seed(n)"=if_else(Survival==0, NA, seed_count_total), # Checked this is always either 0 or NA if live_harvest=="N"
         "Biomass"=if_else(Survival==0, NA, veg_mass),
         "Tiller(n)"=if_else(Survival==0, NA, tillers),
         "Inflorescence(m)"=if_else(Survival==0, NA, inflor_mass),
         "Seed(m)"=`Inflorescence(m)`/`Seed(n)` * 1000, # Megan confirmed this is correct way to calculate Seed mass (1000 miligrams to grams)
         "Seed(m)"=if_else(`Seed(m)`==0, NA, `Seed(m)`), # Two obs at SS where Seed(n)>0 but inflor_mass=0 
         site_numeric=as.numeric(as.factor(site)),
         density_numeric=as.numeric(as.factor(density))-1,
         gravel_numeric=as.numeric(as.factor(albedo))-1,
         year_numeric=as.numeric(as.factor(year))-1,
         environment=site_numeric+density_numeric*4+gravel_numeric*8,
         Biomass=Biomass * 1000) %>% # Need switch from grams to miligrams to match growth chamber data. 
  group_by(genotype, environment) %>%
  summarise(across("Survival":"Seed(m)", ~mean(.x, na.rm=TRUE))) %>%
  dplyr::select(genotype,
                environment,
                "FT",
                "Seed(n)",
                "Seed(m)",
                "Biomass",
                "Tiller(n)",
                "Survival") %>%
  mutate(`Seed(n)`=log(`Seed(n)`),
          Biomass=log(Biomass),
         `Tiller(n)`=log(`Tiller(n)`)) %>%  ### Log transform right skewed distributions
  left_join(kinshipIDs, by="genotype") %>%
  filter(!is.na(NewSiteCode))  %>%
  ungroup()

### Function to calculate standard error
sem <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}

### Check that all traits at all sites are bounded away (at least 2 SE) from 0
allsites_flowered %>%
  group_by(environment) %>%
  summarise(
    across(
      FT:Survival,
      list(
        mean = ~mean(.x, na.rm = TRUE),
        sem  = ~sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x))),
        se_from_0 = ~mean(.x, na.rm = TRUE) /
                    (sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x))))
      ),
      .names = "{.col}_{.fn}"
    )
  )
  
Environments_classification <-
  allsites_flower_harvest %>%
  mutate(site_numeric=as.numeric(as.factor(site)),
         density_numeric=as.numeric(as.factor(density))-1,
         gravel_numeric=as.numeric(as.factor(albedo))-1,
         year_numeric=as.numeric(as.factor(year))-1,
         environment=site_numeric+density_numeric*4+gravel_numeric*8) %>%
  dplyr::select(site, density, albedo, environment) %>% 
  distinct() %>%
  arrange(environment)

### Bring in Physiology Data 
### This has six traits from physiology study measured across 27 genotypes in 2022-2023 in two environments 
genotype_codes <- read.csv(file=file.path(dir, "Data", "BRTEcg_genotypesCode.csv"))
Physiology <- 
  read.csv(file=file.path(dir, "Data", "WC_BZ_2022_2023.csv")) %>%
  ### Common garden data is currently only 2022
  filter(Year==2022) %>%
  ### Match to nomenclature in "Environments_classification"
  mutate(albedo=if_else(Gravel.color=="White", "white", "black"),
         density=if_else(Density=="Low", "lo", "hi"),
         site=if_else(Site=="Balzor", "BA", "WI")) %>%
  ### Joint "Environments_classification" for environment number 
  left_join(Environments_classification, by=c("albedo", "site", "density")) %>% 
  ### Add new site code identifier for genotypes 
  left_join(genotype_codes, by=c("Genotype"="genotype")) %>%
  ### Missing NewSiteCode for genotype 11, currently we'll just remove it
  filter(!is.na(NewSiteCode)) %>%
  ### Select columns we want 
  dplyr::select(environment, NewSiteCode, 
                "Phenological"=Phen_score, 
                "Freeze"=Freezing_Point,
                "Fv/Fm"=dFvFm,
                "SLA"=SLA_m2_kg) %>%
  ### Get genotype means 
  group_by(environment, NewSiteCode) %>%
  summarise(across("Phenological":"SLA", ~mean(.x, na.rm=TRUE))) %>%
  ungroup() %>%
  mutate(across("Phenological":"SLA", ~scale(.x)))

### Kinship matrix (filter for genotypes just observed in common gardens)
genotypes_Physiology=sort(unique(Physiology$NewSiteCode))

### Extract number of traits from Physiology
p_D <- 
  Physiology %>%
  dplyr::select("Phenological":"SLA") %>%
  ncol()
traits_physiology_study <- c("Phenological", "Freeze", "Fv/Fm", "SLA")

###
### Extract all traits and format as array with rows=genotypes, columns=traits, environments=slices 
###

n_g_CG=length(genotypes_common_gardens) # genotypes 
traits_common_garden <- c("FT", "Seed(n)", "Seed(m)", "Biomass", "Tiller(n)", "Survival")
p_CG=length(traits_common_garden) # traits
n=length(unique(allsites_flower_harvest$site))*length(unique(allsites_flower_harvest$albedo))*length(unique(allsites_flower_harvest$density)) # environments

Y_common_gardens_tmp <- vector("list", length=n)
for(i in 1:n){
  Y_mat_CG=matrix(NA, n_g_CG, p_CG)
  Y_mat_D=matrix(NA, n_g_CG, p_D)
  for(j in 1:n_g_CG){
    tmp=allsites_flowered %>%
          filter(environment==i,
                 NewSiteCode==genotypes_common_gardens[j]) %>%
          dplyr::select(`FT`:`Survival`) %>%
          as.matrix() %>%
          c()
    if(length(tmp)>0){
      Y_mat_CG[j,] <-
        allsites_flowered %>%
        filter(environment==i,
               NewSiteCode==genotypes_common_gardens[j]) %>%
        dplyr::select(`FT`:`Survival`) %>%
        as.matrix() %>%
        c()
    }
    if(i %in% unique(Physiology$environment) & 
       genotypes_common_gardens[j] %in% unique(filter(Physiology, environment==i)$NewSiteCode)){
      Y_mat_D[j,] <-
        Physiology %>%
        filter(environment==i,
               NewSiteCode==genotypes_common_gardens[j]) %>%
        dplyr::select(`Phenological`:`SLA`) %>%
        as.matrix() %>%
        c()
    }
  }
  if(i %in% unique(Physiology$environment)){
    Y_common_gardens_tmp[[i]]=as.matrix(cbind(Y_mat_CG, Y_mat_D))
  } else {
    Y_common_gardens_tmp[[i]]=as.matrix(Y_mat_CG)
  }
}

###
### Combine data sources, treat growth_chamber data as one more environment
###

### Joint scaling for all traits common across data sources 
Y_joint <-
  do.call(rbind, map(Y_common_gardens_tmp, ~.x[,1:p_CG])) %>%
  rbind(Y_growth_chamber[,1:p_CG]) %>%
  scale() 

### Create list for joint LVM model 
Y = vector("list", n+1)
Y_common_gardens = vector("list", n)
### Fill in common gardens 
for(i in 1:n){
  idx = ((i-1) * n_g_CG + 1):(n_g_CG * i)
  if(i %in% c(8, 13)){
    Y[[i]] = Y_common_gardens[[i]] = cbind(Y_joint[idx,], Y_common_gardens_tmp[[i]][,(p_CG+1):(p_CG+p_D)])
    colnames(Y[[i]]) = c(traits_common_garden, traits_physiology_study)
    
  } else {
    Y[[i]] = Y_common_gardens[[i]] = cbind(Y_joint[idx,]) 
    colnames(Y[[i]]) = traits_common_garden
  }
}

### Add growth_chamber 
n_g = nrow(genotypes_all)
idx = ((n) * n_g_CG + 1):((n) * n_g_CG + n_g)
Y[[n+1]] = cbind(Y_joint[idx,], scale(Y_growth_chamber[,-c(1:p_CG)]))

### Different W all environments 
W_common_gardens=vector("list", n)
for(i in 1:n){
### There were only 15 genotypes that flowered in Baltzor in white gravel and low density. 
### As a result, there is not enough data to calculate cross correlations between CG and D traits 
### This fills in those cross correlations with what was observed at Wildcat (environment 8)
if(i==13){ 
  Y_cov <-
    Y[[8]] %>%
    cor(use="pairwise.complete.obs") 
  Y_cov13 <-
    Y[[i]] %>%
    cor(use="pairwise.complete.obs") 
  Y_cov[1:p_CG,1:p_CG] <- Y_cov13[1:p_CG,1:p_CG]
  Y_cov[(p_CG+1):(p_CG+p_D),(p_CG+1):(p_CG+p_D)] <- Y_cov13[(p_CG+1):(p_CG+p_D),(p_CG+1):(p_CG+p_D)]
} else {
   Y_cov <-
    Y[[i]] %>%
    cor(use="pairwise.complete.obs") 
}
Y_eigen_cov <- 
  Y_cov %>%
  eigen()
### Find number of principal components needed to capture 90% of variation 
var=cumsum(Y_eigen_cov$values) / sum(Y_eigen_cov$values)
q=which(var>0.90)[1] ### number of principal components 
U_common_gardens=Y_eigen_cov$vectors[,1:q]
Lambda_common_gardens=Y_eigen_cov$values[1:q]
W_common_gardens[[i]]=U_common_gardens%*%diag(sqrt(Lambda_common_gardens))
}

### Find Maximum likelihood estimate for W (see quation (4), Wang 2010)
Y_eigen_cov <- 
  Y[[n+1]] %>%
  cor(use="pairwise.complete.obs") %>%
  eigen()
### Find number of principal components needed to capture 90% of variation 
var=cumsum(Y_eigen_cov$values) / sum(Y_eigen_cov$values)
q=which(var>0.90)[1] ### number of principal components 
### Find nu
U_growth_chamber=Y_eigen_cov$vectors[,1:q]
Lambda_growth_chamber=Y_eigen_cov$values[1:q]
W_growth_chamber=U_growth_chamber%*%diag(sqrt(Lambda_growth_chamber))

### Bind together Ws (different by environment)
W=c(W_common_gardens, list(W_growth_chamber))


### Bring in bioclimate variables 
bioclim <- read.csv(file.path(dir, "Data", "BioclimateOfOrigin_AllGenotypes.csv"))

# Get idx of sites 
idx_sites=which(bioclim$site_code %in% c("BA", "CH", "SS", "WI"))

# Center and scale bioclimatic covariates
bioclim[,6:ncol(bioclim)]=apply(bioclim[,6:ncol(bioclim)], 2, scale)

# Run PCA
pca_out=prcomp(bioclim[,6:ncol(bioclim)])

### Joint model covariate and priors 
X_common_gardens <-
  Environments_classification %>%
  mutate(intercept=1,
         density_n=as.numeric(density=="hi"), # These effects aren't estimated well in the data, huge variance 
         gravel_n=as.numeric(albedo=="black"),
         interaction=gravel_n*density_n,
         PC1=rep(pca_out$x[idx_sites, 1], 4),
         PC2=rep(pca_out$x[idx_sites, 2], 4)) %>%
  dplyr::select("intercept":"PC2") %>%
  as.matrix()
n=length(Y)
X_growth_chamber=matrix(0,1,ncol(X_common_gardens))
X=rbind(X_common_gardens, X_growth_chamber)
d=ncol(X)
z=c(rep(1,nrow(X)-1),0)
X=cbind(
  c(rep(0,n-1), 1), ### add column for growth_chamber mean 
  X
)
d=ncol(X)
X_list=vector("list", n)
for(i in 1:n){
  p=ncol(Y[[i]])
  n_g=nrow(Y[[i]])
  X_unique=matrix(NA, n_g, d)
  for(l in 1:d){
    X_unique[,l]=rep(X[i,l], each=n_g)
  }
  X_list[[i]]=X_unique
}

### Get list of missing observations 
NA_idx=vector("list", n)
for(i in 1:n){
  NA_idx[[i]]=which(is.na(Y[[i]]), arr.ind=TRUE)-1
}

### Hyperparameters 
q_K=q_e=0.001
r_K=r_e=1/q_e
mu_beta=rep(0, d)
Sigma_beta=diag(100^2, d)
mu_0=0
sigma2_0=100^2
scale=1 
scale_zeta2=1 

### Kinship matrix (filter for genotypes just observed in common gardens)
K_common_gardens=K_joint[genotypes_common_gardens,genotypes_common_gardens]
K_list=vector("list", 2)
K_list[[1]]=K_common_gardens
K_list[[2]]=K_joint
### Filter common gardens array for observed genentypes 
genotypes_common_gardens_idx=which(colnames(K_joint) %in% genotypes_common_gardens)

### MCMC iterations 
n_mcmc=20000

### Indexes for trait correlations 
p_idx=vector("list", length(Y))
### label traits 0-5 for 6 common garden traits, 6-9 for Physiology traits, and 10:13 growth_chamber traits
for(i in 1:n){
  p_idx[[i]]=1:ncol(Y[[i]])-1 
}
p_idx[[length(Y)]]=c(1:6,11:14)-1
p_tot=14
### Indexes for what traits in which environments 
n_idx=replicate(p_tot, 1:17-1, simplify = FALSE)
n_idx[6:10]=replicate(length(7:11), c(8,13)-1, simplify = FALSE) # Physiology traits only in two common gardens environments 
n_idx[11:p_tot]=replicate(length(12:p_tot), 17-1, simplify = FALSE) # The remainder of traits are only in growth_chamber 

### Extract missing data in each environment 
n=length(Y)
NA_idx=vector("list", n)
for(i in 1:n){
  NA_idx[[i]]=which(is.na(Y[[i]]), arr.ind=TRUE)-1
}

fix_signs <- function(loadings) {
  # loadings: a matrix like prcomp$rotation or eig$vectors
  fixed <- loadings
  for (j in 1:ncol(loadings)) {
    # find the variable with largest absolute loading in this PC
    max_idx <- which.max(abs(loadings[, j]))
    # if that loading is negative, flip the whole column
    if (loadings[max_idx, j] < 0) {
      fixed[, j] <- -fixed[, j]
    }
  }
  fixed
}

### Save as output list 
data_list <- 
  list(Y=Y, Physiology=Physiology, W=W, X=X,
       K_list=K_list, X_list=X_list, NA_idx=NA_idx,
       q_K=q_K, r_K=r_K, q_e=q_e, r_e=r_e, 
       mu_0=mu_0, sigma2_0=sigma2_0, scale=scale, scale_zeta2=scale_zeta2,
       Environments_classification=Environments_classification,
       mu_beta=mu_beta,Sigma_beta=Sigma_beta, Ancestry=Ancestry,
       p_idx=p_idx, p_tot=p_tot, n_idx=n_idx)


### Save list 
save(data_list, file=file.path(dir, "Outputs", "data_list.RData"))
```



# Parallel Simulation Study 

## Simulate Datasets 

```{r, Parallel Simulation Study Heritability Estimation}

###
### Simulate Datasets 
###

set.seed(3235)

# Define the simulation settings
sim_settings <- list(
  list(G = 50, T = 5,  E = 5),  # many genotypes 
  list(G = 25, T = 10, E = 5), # many traits 
  list(G = 25, T = 5,  E = 10)  # many environments 
)

# Number of datasets per setting
n_sim <- 50

# Function to simulate multiple datasets per setting
simulate_multiple <- function(G, T, E, n_sim, P = 1, percent_remove = 0.1) {
  map(1:n_sim, ~ simulate_data(G = G, T = T, E = E, P = P, percent_remove = percent_remove, dir = dir))
}

# Create datasets for all settings
Start=Sys.time()
datasets <- map(sim_settings, ~simulate_multiple(.x$G, .x$T, .x$E, n_sim)) %>%
  flatten()  # <-- flatten one level
End=Sys.time()
difftime(End,Start) # ~11 seconds for n_sim=50

# save 
save(datasets, file=file.path(dir, "Outputs", "datasets.RData"))
```

## Run Simulation Study in Parallel 

```{r, Fit Models to Simulated Data}
# load datasets 
load(file=file.path(dir, "Outputs", "datasets.RData"))

# Set up parallel plan
plan(multisession, workers = 10)

# "datasets" is a flat list of simulated datasets
Start <- Sys.time()

safe_simulation_study <- safely(simulation_study)

results <- future_map(
  datasets,
  ~safe_simulation_study(data = .x),
  .progress = TRUE,
  models = c("KO", "MMM", "UO", "CO", "CCO"),               
  chains = 1, parallel_chains = 1,
  iter_warmup = 200, iter_sampling = 1000,
  refresh = 100,
  seed = TRUE
)

# Separate successes from errors
metrics_out <- map_dfr(results, "result", .id = "dataset_id")
errors <- map(results, "error")


End <- Sys.time()
difftime(End, Start) # ~ 7 hours with 10 workers 

# save 
save(metrics_out, file=file.path(dir, "Outputs", "metrics_out.RData"))
```

## Make Simulation Study Plots 

```{r, Simulation Study Plots}
load(file=file.path(dir, "Outputs", "metrics_out.RData"))

# --- Create dataset column as proper expressions ---
metrics_long <- metrics_out %>%
  mutate(
    esr = log10( ess / time ),
    esr_b =  log10( ess_b / time ) ,
    esr_Y =  log10( ess_Y / time )
  ) %>%
  pivot_longer(
    cols = c(
      coverage, bias, sd, cor, esr, 
      coverage_b, bias_b, sd_b, cor_b, esr_b, 
      coverage_Y, bias_Y, sd_Y, cor_Y, esr_Y),
    names_to = "metric",
    values_to = "value"
  )  %>%
  mutate(
    goal = case_when(
      str_detect(metric, "_Y$")  ~ "Posterior Prediction",
      str_detect(metric, "_b$")  ~ "Beta Estimation",
      TRUE                        ~ "Heritability Estimation"
    ),
    # remove the suffixes for clean metric names
    metric = str_remove(metric, "_Y$|_b$"),
    metric = factor(
      metric, 
      levels = c("esr", "coverage", "bias", "sd", "cor"),
      labels = c("Efficiency", "Coverage", "Bias", "Variability", "Correlation")
    ),
    goal = factor(goal, levels = c("Heritability Estimation", "Posterior Prediction", "Beta Estimation")),
    model = factor(model, levels = c("MMM", "KO", "CCO", "CO", "UO")),
    dataset = case_when(
      G == 50 ~ "G==50",
      T == 10 ~ "T==10",
      E == 10 ~ "E==10",
      TRUE ~ "Other"
    ),
    dataset = factor(dataset, levels = c("G==50", "T==10", "E==10"))
  ) %>%
  filter(
    metric != "Correlation", # This metric doesn't add much and is a lot to plot 
    !(model %in% c("CCO", "UO", "CO") & goal == "Heritability Estimation"), # Remove metrics that certain models can't do 
    !(model %in% c("UO") & goal == "Beta Estimation")
    )

# --- Split by goal --- 
metrics_H2 <- metrics_long %>% filter(goal == "Heritability Estimation") 
metrics_beta <- metrics_long %>% filter(goal == "Beta Estimation") 
metrics_pred <- metrics_long %>% filter(goal == "Posterior Prediction")

# --- Boxplot function ---
plot_box <- function(df, title) {
  ggplot(df, aes(x = model, y = value, fill = model)) +
    geom_boxplot(outlier.shape = NA) +   # Remove outliers
    facet_grid(metric ~ dataset, scales = "free", labeller = label_parsed) +
    scale_fill_manual(values = wes_palette("Darjeeling1"),
                      labels = 
                        c("Multivariate Mixed Model", 
                          "Kinship Ordination", 
                          "Concurrent Ordination",
                          "Constrained Ordination",
                          "Unconstrained Ordination")) +
    theme_bw(base_size = 14) +
    labs(
      x = "Model",
      y = "Value",
      fill = "Model",
      title = title
    ) +
    theme(
      strip.background = element_rect(fill = "grey90", color = NA),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom"
    )
}

# --- Plot ---
p_H2 <- plot_box(metrics_H2, "Heritability Estimation")
p_beta <- plot_box(metrics_beta, "Beta Estimation")
p_pred <- plot_box(metrics_pred, "Posterior Prediction")


p_H2
p_beta
p_pred

# --- Save ---
ggsave(plot=p_H2, 
       filename=file.path(dir, "Figures", "Heritability_boxblot.png"),  
       dpi=1000, height=7, width=7)

ggsave(plot=p_beta, 
       filename=file.path(dir, "Figures", "beta_boxblot.png"),  
       dpi=1000, height=7, width=10)

ggsave(plot=p_pred, 
       filename=file.path(dir, "Figures", "Prediction_boxblot.png"),  
       dpi=1000, height=7, width=11)
```

# Real Data Analysis (Cheatgrass: Growth Chamber + Common Gardens + Physiology)

## Data Preparation (need to run for all scripts below)

```{r, Real Data Analysis (Cheatgrass: Growth Chamber + Common Gardens + Physiology)}

### Load Data 
load(file=file.path(dir, "Outputs", "data_list.RData"))

###
###
### Reformat data for model fitting with stan
###
###

### Full kinship matrix  
K <- data_list$K_list[[2]]
genotypes <- colnames(K)

###
### Common Garden
###

### Dimensions 
E <- 16 
traits_common_garden <- c("FT", "Seed(n)", "Seed(m)", "Biomass", "Tiller(n)", "Survival")
T <- length(traits_common_garden)
G <- nrow(data_list$Y[[1]])

### Data
data_common_garden <-
  map(1:E, ~ data_list$Y[[.x]][, 1:6]) %>%
  abind(along = 1) %>%
  # Traits have been scaled to have mean 0 variance 1 across all environments 
  # This includes the growth chamber that we are excluding here so need to do rescale 
  # Silence this for creating data all because then we want unscaled 
  #scale() %>% 
  `colnames<-`(traits_common_garden) %>%
  as.data.frame() %>%
  mutate(environment = as.factor(rep(1L:E, each = G)),
         genotype = rep(colnames(data_list$K_list[[1]]), E),
         genotype = factor(genotype, levels = genotypes)) %>%
  gather(key="trait", value = "value", -c("genotype", "environment")) %>%
  mutate(trait = factor(trait, levels=traits_common_garden, labels=traits_common_garden)) %>%
  # Want all unobserved levels to get included as NA for posterior prediction 
  complete(
    genotype = levels(genotype),
    trait = levels(trait),
    environment = levels(environment),
    fill = list(value = NA)
  ) %>%
  mutate(
    g = as.integer(factor(genotype, levels = genotypes)),
    t = as.integer(factor(trait, levels = traits_common_garden)),
    e = as.integer(factor(environment, levels = 1:E))
  ) %>%
  arrange(e, t, g)
        
### Get rows with NAs 
NAs <- 
  data_common_garden$value %>%
  is.na() %>%
  which()

### Build the list
list_common_garden <- list(
  # Dimensions 
  N = length(data_common_garden$value[-NAs]),
  N_p = length(data_common_garden$value[NAs]),
  P = ncol(data_list$X[-17,-1]),
  G = max(data_common_garden$g),
  E = max(data_common_garden$e),
  T = max(data_common_garden$t),
  T_a = max(data_common_garden$t),
  q = 4, 
  # In Sample 
  g_idx = data_common_garden$g[-NAs], 
  e_idx = data_common_garden$e[-NAs],   
  t_idx = data_common_garden$t[-NAs],   
  y = data_common_garden$value[-NAs],
  X = data_list$X[-17,-1],
  K = K,
  # Out of sample
  g_idx_p = data_common_garden$g[NAs], 
  e_idx_p = data_common_garden$e[NAs],   
  t_idx_p = data_common_garden$t[NAs]  
)

###
### Physiology Study
###

### dimensions  
traits_physiology_study <- c("Phenological", "Freeze", "Fv/Fm",  "SLA")
physiology_genotypes <- unique(data_list$Physiology$NewSiteCode)

### Data
data_physiology <-
  data_list$Physiology %>%
  as.data.frame() %>%
  mutate(environment = factor(environment, levels = c(8, 13)),
         genotype = factor(NewSiteCode, levels = genotypes)) %>%
  dplyr::select(-c("NewSiteCode")) %>%
  gather(key="trait", value = "value", -c("genotype", "environment")) %>%
  mutate(trait = factor(trait, levels=traits_physiology_study, labels=traits_physiology_study)) %>%
  # Want all unobserved levels to get included as NA for posterior prediction 
  complete(
    genotype = levels(genotype),
    trait = levels(trait),
    environment = levels(environment),
    fill = list(value = NA)
  ) %>%
  mutate(
    g = as.integer(factor(genotype, levels = genotypes)),
    t = as.integer(factor(trait, levels = traits_physiology_study)),
    e = as.integer(factor(environment))
  ) %>%
  arrange(e, t, g)
        
### Get rows with NAs 
NAs <- 
  data_physiology$value %>%
  is.na() %>%
  which()

### Build the list
list_physiology_study <- list(
  # Dimensions 
  N = length(data_physiology$value[-NAs]),
  N_p = length(data_physiology$value[NAs]),
  P = 2,
  G = max(data_physiology$g),
  E = max(data_physiology$e),
  T = max(data_physiology$t),
  T_a = 0, # too few environments to justify pooling 
  q = 2, 
  # In Sample 
  g_idx = data_physiology$g[-NAs], 
  e_idx = data_physiology$e[-NAs],   
  t_idx = data_physiology$t[-NAs],   
  y = data_physiology$value[-NAs],
  X = diag(1, 2),
  K = K,
  # Out of sample
  g_idx_p = data_physiology$g[NAs], 
  e_idx_p = data_physiology$e[NAs],   
  t_idx_p = data_physiology$t[NAs]  
)

###
### Growth Chamber
###

### Dimensions 
traits_growth_chamber <- 
  c("FT", "Seed(n)", "Seed(m)", "Biomass", "Tiller(n)",  
    "Survival","Height", "Growth", "Leaves(n)",  "Inflorescence(h)")

data_growth_chamber <-
  data_list$Y[[17]] %>%
  # Traits have been scaled to have mean 0 variance 1 across all environments 
  # This includes the common gardebs that we are excluding here so need to do rescale 
  # Silence if creating data_all
  #scale() %>% 
  `colnames<-`(traits_growth_chamber) %>%
  as.data.frame() %>%
  mutate(environment = factor(c(17)),
         genotype = factor(genotypes, levels = genotypes)) %>%
  gather(key="trait", value = "value", -c("genotype", "environment")) %>%
  mutate(trait = factor(trait, levels=traits_growth_chamber, labels=traits_growth_chamber)) %>%
  # This does nothing because all levels already there but include for completeness 
  complete(
    genotype = levels(genotype),
    trait = levels(trait),
    environment = levels(environment),
    fill = list(value = NA)
  ) %>%
  mutate(
    g = as.integer(factor(genotype, levels = genotypes)),
    t = as.integer(factor(trait, levels = traits_growth_chamber)),
    e = as.integer(factor(environment))
  ) %>%
  arrange(e, t, g)

### Get rows with NAs 
NAs <- 
  data_growth_chamber$value %>%
  is.na() %>%
  which()

### Build the list
list_growth_chamber <- list(
  # Dimensions 
  N = length(data_growth_chamber$value[-NAs]),
  N_p = length(data_growth_chamber$value[NAs]),
  P = 1,
  G = max(data_growth_chamber$g),
  E = max(data_growth_chamber$e),
  T = max(data_growth_chamber$t),
  T_a = 0, # only 1 environment, so no pooling
  q = 4,
  # In Sample 
  g_idx = data_growth_chamber$g[-NAs], 
  e_idx = data_growth_chamber$e[-NAs],   
  t_idx = data_growth_chamber$t[-NAs],   
  y = data_growth_chamber$value[-NAs],
  X = diag(1, 1),
  K = K,
  # Out of sample
  g_idx_p = data_growth_chamber$g[NAs], 
  e_idx_p = data_growth_chamber$e[NAs],   
  t_idx_p = data_growth_chamber$t[NAs]  
)

###
### Combined Datasets and List 
###

traits_all <-
  c(
    traits_common_garden, 
    traits_physiology_study, 
    traits_growth_chamber
    ) %>%
  unique()

data_all <-
  rbind(
    data_common_garden,
    data_physiology,
    data_growth_chamber
  ) %>%
  mutate(trait = factor(trait, 
                        levels = traits_all, 
                        labels = traits_all),
         t = as.integer(trait),
         e = as.integer(environment),
         g = as.integer(factor(genotype, levels = genotypes))) %>%
  arrange(e, t, g)


### Get rows with NAs 
NAs <- 
  data_all$value %>%
  is.na() %>%
  which()

### Build the list
list_all <- list(
  # Dimensions 
  N = length(data_all$value[-NAs]),
  N_p = length(data_all$value[NAs]),
  P = ncol(data_list$X),
  G = max(data_all$g),
  E = max(data_all$e),
  T = max(data_all$t),
  T_a = 6,
  q = 4,
  # In Sample 
  g_idx = data_all$g[-NAs], 
  e_idx = data_all$e[-NAs],   
  t_idx = data_all$t[-NAs],   
  y = data_all$value[-NAs],
  X = data_list$X,
  K = K,
  # Out of sample
  g_idx_p = data_all$g[NAs], 
  e_idx_p = data_all$e[NAs],   
  t_idx_p = data_all$t[NAs]
)
```

## Fit Models  

```{r, Fit Models Bromus tectorum}
            
### Hyperparameters 
chains <- 3
parallel_chains <- 3
iter_warmup <- 200
iter_sampling <- 1000
refresh <- 100

###
### Fit Kinship Ordination  
###

KO_fit <- 
  fit_model(
    list_all,
    dir = dir,
    chains, iter_warmup, iter_sampling, refresh,
    model = c("KO"),
    pathfinder = TRUE
  )

### Check convergence of identifiable quantities 
hist(KO_fit$summary("beta")$rhat)
hist(KO_fit$summary("H2")$rhat)
hist(KO_fit$summary("sigma2")$rhat)
hist(KO_fit$summary("sigma2_sigma2")$rhat)
hist(KO_fit$summary("mu_sigma2")$rhat)
hist(KO_fit$summary("y_pred")$rhat)

### not identifiable, but might be stable just by chance 
hist(KO_fit$summary("W")$rhat)
hist(KO_fit$summary("zeta2")$rhat)
hist(KO_fit$summary("eta")$rhat)

### Load model if already fit
# saveRDS(KO_fit, file.path(dir, "Outputs", "KO_fit_q4.rds"))
# KO_fit <- readRDS(file.path(dir, "Outputs", "KO_fit_q4.rds"))
            

###
### Fit Multivariate Mixed Model to experiments independently  
###

### Common gardens 
MMM_fit_CG <- 
  fit_model(
    list_common_garden,
    dir = dir,
    chains, iter_warmup, iter_sampling, refresh,
    model = c("MMM"),
    pathfinder = TRUE
  )

### Check convergence of identifiable quantities 
hist(MMM_fit_CG$summary("beta")$rhat)
hist(MMM_fit_CG$summary("H2")$rhat)
hist(MMM_fit_CG$summary("zeta2")$rhat)
hist(MMM_fit_CG$summary("sigma2")$rhat)
hist(MMM_fit_CG$summary("sigma2_sigma2")$rhat)
hist(MMM_fit_CG$summary("mu_sigma2")$rhat)


### Save fit
# saveRDS(MMM_fit_CG, file.path(dir, "Outputs", "MMM_fit_CG.rds"))
# readRDS(file.path(dir, "Outputs", "MMM_fit_CG.rds"))

### Growth Chamber
MMM_fit_GC <- 
  fit_model(
    list_growth_chamber,
    dir = dir,
    chains, iter_warmup, iter_sampling, refresh,
    model = c("MMM"),
    pathfinder = FALSE
  )

### Check convergence of identifiable quantities 
hist(MMM_fit_GC$summary("beta")$rhat)
hist(MMM_fit_GC$summary("H2")$rhat)
hist(MMM_fit_GC$summary("sigma2")$rhat)
hist(MMM_fit_GC$summary("zeta2")$rhat)

### Save fit
# saveRDS(MMM_fit_GC, file.path(dir, "Outputs", "MMM_fit_GC.rds"))
# readRDS(file.path(dir, "Outputs", "MMM_fit_GC.rds"))

### Physiology study 
MMM_fit_PS <- 
  fit_model(
    list_physiology_study,
    dir = dir,
    chains, iter_warmup, iter_sampling, refresh,
    model = c("MMM"),
    pathfinder = TRUE
  )

### Check convergence of identifiable quantities 
hist(MMM_fit_PS$summary("beta")$rhat)
hist(MMM_fit_PS$summary("H2")$rhat)
hist(MMM_fit_PS$summary("sigma2")$rhat)
hist(MMM_fit_PS$summary("zeta2")$rhat)

### Save fit
# saveRDS(MMM_fit_PS, file.path(dir, "Outputs", "MMM_fit_PS.rds"))
# readRDS(file.path(dir, "Outputs", "MMM_fit_PS.rds"))

```

## Heritability Plot 
   
```{r, Heritability Plot }     

###
### Load Model Fits & Metadata
###

### Model Fits 
KO_fit <- readRDS(file.path(dir, "Outputs", "KO_fit_q4.rds"))
MMM_fit_CG <- readRDS(file.path(dir, "Outputs", "MMM_fit_CG.rds"))
MMM_fit_GC <- readRDS(file.path(dir, "Outputs", "MMM_fit_GC.rds"))
MMM_fit_PS <- readRDS(file.path(dir, "Outputs", "MMM_fit_PS.rds"))

### Metadata 
load(file=file.path(dir, "Outputs", "data_list.RData"))

### Bring in data on environments 
Environments_classification <- 
  data_list$Environments_classification %>%
  mutate(density=if_else(density=="hi", "High", "Low"),
         gravel=if_else(albedo=="black", "Black", "White"),
         treatment=paste(density, gravel, sep=" "),
         site=case_when(site=="BA" ~ "Baltzor",
                        site=="CH" ~ "Cheyenne",
                        site=="SS" ~ "Sheep Station",
                        site=="WI" ~ "Wildcat"),
         environment=environment) %>%
  mutate(
    environment_text = paste(site, treatment, sep=", "),
    environment_text = factor(environment_text, levels = c(
    "Sheep Station, High White", "Sheep Station, Low White", "Sheep Station, High Black", "Sheep Station, Low Black",
    "Cheyenne, High White", "Cheyenne, Low White", "Cheyenne, High Black", "Cheyenne, Low Black",
    "Baltzor, High White", "Baltzor, Low White", "Baltzor, High Black", "Baltzor, Low Black",
    "Wildcat, High White", "Wildcat, Low White", "Wildcat, High Black", "Wildcat, Low Black"
  ))) %>%
  rbind(
    tibble(
      site = "Growth Chamber",
      density = NA,
      albedo = NA,
      environment = 17L,
      gravel = NA,
      treatment = NA,
      environment_text = factor("Growth Chamber", levels=c("Growth Chamber"))
    )
  ) %>%
  mutate(environment=as.character(environment))


###
### Create tibbles 
###

### Kinship ordination 
H2_KO <-
  KO_fit$summary("H2") %>%
  # Extract trait and environment indices from the "variable" column
  extract(variable, into = c("trait", "environment"), regex = "H2\\[(\\d+),(\\d+)\\]", convert = TRUE) %>%
  # Optionally, attach labels if you want actual trait/environment names
  mutate(
    Model="KO",
    experiment = case_when(
      trait %in% 1:6 & environment <= 16  ~ "common garden",
      trait %in% 7:10 & environment %in% c(8, 13) ~ "physiology study",
      trait %in% c(1:6, 11:14) & environment > 16 ~ "growth chamber"
    ),
    trait = levels(data_all$trait)[trait],
    trait = factor(trait, levels=traits_all), # for ordering in plots 
    environment = unique(data_all$environment)[environment],
    CI_width = q95 - q5
  ) %>%
  left_join(
    Environments_classification, by="environment"
  ) %>%
  # Filter for observed trait, environment combinations (i.e, trait "Freeze" not measured in environment 1)
  filter(!is.na(experiment))

###
### Multivariate Mixed Model 
###

### Commom garden 
H2_MMM_CG <-
  MMM_fit_CG$summary("H2") %>%
  # Extract trait and environment indices from the "variable" column
  extract(variable, into = c("trait", "environment"), regex = "H2\\[(\\d+),(\\d+)\\]", convert = TRUE) %>%
  # Optionally, attach labels if you want actual trait/environment names
  mutate(
    Model="MMM",
    experiment = "common garden",
    trait = traits_common_garden[trait],
    trait = factor(trait, levels=traits_common_garden), # for ordering in plots 
    environment = unique(data_common_garden$environment)[environment],
    CI_width = q95 - q5
  ) %>%
  left_join(
    Environments_classification, by="environment"
  ) 

### Physiology study 
H2_MMM_PS <-
  MMM_fit_PS$summary("H2") %>%
  # Extract trait and environment indices from the "variable" column
  extract(variable, into = c("trait", "environment"), regex = "H2\\[(\\d+),(\\d+)\\]", convert = TRUE) %>%
  # Optionally, attach labels if you want actual trait/environment names
  mutate(
    Model="MMM",
    experiment = "physiology study",
    trait = traits_physiology_study[trait],
    trait = factor(trait, levels=traits_physiology_study), # for ordering in plots 
    environment = rep(c("8", "13"), each=length(traits_physiology_study)),
    CI_width = q95 - q5
  ) %>%
  left_join(
    Environments_classification, by="environment"
  ) 

### Growth chamber 
H2_MMM_GC <-
  MMM_fit_GC$summary("H2") %>%
  # Extract trait and environment indices from the "variable" column
  extract(variable, into = c("trait", "environment"), regex = "H2\\[(\\d+),(\\d+)\\]", convert = TRUE) %>%
  # Optionally, attach labels if you want actual trait/environment names
  mutate(
    Model="MMM",
    experiment = "growth chamber",
    trait = traits_growth_chamber[trait],
    trait = factor(trait, levels=traits_growth_chamber), # for ordering in plots 
    environment = "17",
    CI_width = q95 - q5
  ) %>%
  left_join(
    Environments_classification, by="environment"
  ) 

### Combine 
H2_MMM <-
  rbind(
    H2_MMM_CG,
    H2_MMM_PS,
    H2_MMM_GC
  ) %>%
  # Refactor traits 
  mutate(
    trait = factor(trait, levels = traits_all)
  )

### Percent difference in CI width of methods 
(mean(H2_KO$CI_width) - mean(H2_MMM$CI_width)) / mean(H2_MMM$CI_width)

### Join tibbles 
H2 <- 
  rbind(
    H2_KO,
    H2_MMM
  ) %>%
  mutate(
    environment = as.integer(environment),
    Model = factor(Model, levels = c("MMM", "KO"))
  )

### Check correlation of estimates (within traits)
cor_results <- 
  H2 %>%
  filter(trait %in% traits_common_garden) %>% # Only can do this for traits observed in 3 or more environments 
  dplyr::select(Model, environment, trait, mean) %>%
  pivot_wider(names_from = Model, values_from = mean) %>%
  group_by(trait) %>%
  summarise(cor_mean = cor(MMM, KO, use = "complete.obs"),
            .groups = "drop")

### Average correlation within traits  
mean(cor_results$cor_mean)

###
### Plot Heritability in common gardens 
###

heritability_common_gardens_plot <-
  H2 %>%
  filter(experiment == "common garden") %>%
  ggplot(aes(x=trait, y=mean, ymin=q5, ymax=q95, color=Model)) +
  facet_grid(site~density + gravel, scales = "fixed") +
  geom_point(position = position_dodge(width = 0.9)) +
  geom_linerange(position = position_dodge(width = 0.9)) +
  theme_bw() +
  xlab("Trait") +
  ylab("Heritability") +
  ylim(c(0,1)) +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("Common Gardens") +
  scale_color_manual(name="Model", 
                     labels=c("Multivariate Mixed Model", "Kinship Ordination"),
                     values=wes_palette("Darjeeling1", 2, type = c("discrete")))

### Save plot 
ggsave(plot=heritability_common_gardens_plot, 
       filename=file.path(dir, "Figures", "heritability_common_gardens.png"),  
       dpi=1000, height=8, width=7)

### Check correlation of posterior means across environments and traits 
cor(
  filter(H2, Model=="MMM", experiment=="common garden")$mean,
  filter(H2, Model=="KO", experiment=="common garden")$mean
)

### Percent difference in credible interval widths 
(mean(filter(H2, Model=="KO", experiment=="common garden")$CI_width) - 
 mean(filter(H2, Model=="MMM", experiment=="common garden")$CI_width)) / 
 mean(filter(H2, Model=="MMM", experiment=="common garden")$CI_width)


###
### Physiology Study
###

### Make plot of estimates 
heritability_physiology_plot <-
  H2 %>%
  filter(experiment == "physiology study") %>%
  ggplot(aes(x=trait, y=mean, ymin=q5, ymax=q95, color=Model)) +
  facet_grid(site~density + gravel, scales = "fixed") +
  geom_point(position = position_dodge(width = 0.9)) +
  geom_linerange(position = position_dodge(width = 0.9)) +
  theme_bw() +
  xlab("Trait") +
  ylab("Heritability") +
  ylim(c(0,1)) +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("Physiology Study") +
  scale_color_manual(name="Model", 
                     labels=c("Multivariate Mixed Model", "Kinship Ordination"),
                     values=wes_palette("Darjeeling1", 2, type = c("discrete")))

### Save plot 
ggsave(plot=heritability_physiology_plot, 
       filename=file.path(dir, "Figures", "heritability_physiology_plot.png"),  
       dpi=1000, height=4, width=7)

### Check correlation of posterior means across environments and traits 
cor(
  filter(H2, Model=="MMM", experiment=="physiology study")$mean,
  filter(H2, Model=="KO", experiment=="physiology study")$mean
)

### Percent difference in credible interval widths 
(mean(filter(H2, Model=="KO", experiment=="physiology study")$CI_width) - 
 mean(filter(H2, Model=="MMM", experiment=="physiology study")$CI_width)) / 
 mean(filter(H2, Model=="MMM", experiment=="physiology study")$CI_width)

###
### Growth Chamber
###

### Make plot of estimates 
heritability_growth_chamber_plot <- 
  H2 %>%
  filter(experiment == "growth chamber") %>%
  ggplot(aes(x=trait, y=mean, ymin=q5, ymax=q95, color=Model)) +
  geom_point(position = position_dodge(width = 0.9)) +
  geom_linerange(position = position_dodge(width = 0.9)) +
  theme_bw() +
  xlab("Trait") +
  ylab("Heritability") +
  ylim(c(0,1)) +
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  ggtitle("Growth Chamber") +
  scale_color_manual(name="Model", 
                     labels=c("Multivariate Mixed Model", "Kinship Ordination"),
                     values=wes_palette("Darjeeling1", 2, type = c("discrete")))

### Save plot 
ggsave(plot=heritability_growth_chamber_plot, 
       filename=file.path(dir, "Figures", "heritability_growth_chamber.png"),  
       dpi=1000, height=4, width=7)

### Check correlation of posterior means across environments and traits 
cor(
  filter(H2, Model=="MMM", experiment=="growth chamber")$mean,
  filter(H2, Model=="KO", experiment=="growth chamber")$mean
)

### Percent difference in credible interval widths 
(mean(filter(H2, Model=="KO", experiment=="growth chamber")$CI_width) - 
 mean(filter(H2, Model=="MMM", experiment=="growth chamber")$CI_width)) / 
 mean(filter(H2, Model=="MMM", experiment=="growth chamber")$CI_width)


###
### Combine all heritability estimates in one plot 
###


# Plot with colored facet labels based principal component value 
colors <- c("darkblue", "white", "darkred") # Match colors in other BromeCast manuscripts 
pc1_values = c(-3.2051, -2.0867, 0.3363, 2.3312)
pc1_values_scaled = (pc1_values - min(pc1_values))/sum((pc1_values - min(pc1_values)))+0.25 ### To do scale and center to c(0,1)
env_colors <- scales::gradient_n_pal(colors)
env_colors_labels <- env_colors(pc1_values_scaled)

### Plot (both methods)
heritability_plot <-
  H2 %>%
  ggplot(aes(x=trait, y=mean, ymin=q5, ymax=q95, color=Model)) +
  geom_point(position = position_dodge(width = 0.9)) +
  geom_linerange(position = position_dodge(width = 0.9)) +
  theme_bw() +
  ylab("Narrow-sense Heritability") +
  xlab("") +
  facet_wrap2(
    ~environment_text, 
    scales = "free", 
    strip = strip_themed(
      background_x = elem_list_rect(fill = c(rep(env_colors_labels, each=4), "grey"))
    ),
    ncol=4, 
    nrow=5
  ) +
  theme(legend.position = c(0.6, 0.05), 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        legend.text = element_text(size=12)) +
  scale_color_manual(name="Model", 
                     labels=c("Multivariate Mixed Model", "Kinship Ordination"),
                     values=wes_palette("Darjeeling1", 2, type = c("discrete")))

### Save plot 
ggsave(plot=heritability_plot, 
       filename=file.path("Figures", "heritability_plot.png"),  
       dpi=200, height=10, width=9)
```


## Trait Selection Plot 

```{r, Real Data Analysis (Exlude fitness traits)}

###
### Calculate Trait Selection Coefficients  
###

### Load model fit for PPD samples 
KO_fit <- readRDS(file.path(dir, "Outputs", "KO_fit_q4.rds"))
y_pred <- KO_fit$draws("y_pred")

###
### Dimensions 
###

### mcmc samples
L = dim(y_pred)[1]
### environments 
E = list_all$E

### Create arrays for storing posterior samples 
Y <- array(
  NA,
  dim = c(list_all$G, list_all$T, list_all$E, L)
)

### Add observed traits
CG_genotypes <- colnames(data_list$K_list[[1]]) # Extract common garden genotypes 
data_obs <-
  data_all %>%
  filter(!is.na(value),
         ### We are interested in invasion in the Intermountain West so stick with those genotypes 
         genotype %in% CG_genotypes)  
N <- nrow(data_obs)
for(i in 1:N){
  g = data_obs$g[i]
  t = data_obs$t[i]
  e = data_obs$e[i]
  Y[g,t,e,] = data_obs$value[i]
}

### Add PPD
data_pred <-
  data_all %>%
  filter(is.na(value),
         ### We are interested in invasion in the Intermountain West so stick with those genotypes 
         genotype %in% CG_genotypes)
N_p <- nrow(data_pred)
for(i in 1:N_p){
  g = data_pred$g[i]
  t = data_pred$t[i]
  e = data_pred$e[i]
  Y[g,t,e,] = y_pred[,1,i] ### just use one chain 
}

### Calculate log(fitness) = log(Reproduce) + log(Seed(n))
fitness <- sign(Y[,6,,])*log(abs(Y[,6,,])) + Y[,2,,]

### Store coefficients 
S_KO <- vector("list", E)

for(e in 1:E){
  if(e %in% c(8, 13)){
    traits = c(1, 3, 5, 7:10)
  } else if (e == E){
    # No Biomass or Inflorescence height because that is predicting fitness using pseudo fitness  
    traits = c(1, 3, 5, 11:13)
  } else {
    traits = c(1, 3, 5)
  }
  pred = Y[,traits,e,]
  T = ncol(pred)
  S_mat = matrix(NA, T, L)
  for(l in 1:L){
    S_mat[,l] = lm(fitness[,e,l] ~ pred[,,l])$coefficients[-1]
  }
  S_KO[[e]] = S_mat
}


###
### Extract and plot results 
###

### Variable Names 
common_gardens_traits=c("FT", "Seed(m)", "Tiller(n)")
traits_physiology_study=c("FT", "Seed(m)", "Tiller(n)", "Phenological", "Freeze", "Fv/Fm",  "SLA")
traits_growth_chamber=c("FT", "Seed(m)", "Tiller(n)", "Height", "Growth", "Leaves(n)")

Variable=c(
  rep(common_gardens_traits, 7),
  traits_physiology_study,
  rep(common_gardens_traits, 4),
  traits_physiology_study,
  rep(common_gardens_traits, 3),
  traits_growth_chamber
) 
environment=c(
  rep(1:7, each=length(common_gardens_traits)),
  rep(8, each=length(traits_physiology_study)),
  rep(9:12, each=length(common_gardens_traits)),
  rep(13, each=length(traits_physiology_study)),
  rep(14:16, each=length(common_gardens_traits)),
  rep(17, each=length(traits_growth_chamber))
)

### Create tibbles
Environments_classification <- 
  data_list$Environments_classification %>%
  mutate(density=if_else(density=="hi", "High", "Low"),
         gravel=if_else(albedo=="black", "Black", "White"),
         treatment=paste(density, gravel, sep=" "),
         site=case_when(site=="BA" ~ "Baltzor",
                        site=="CH" ~ "Cheyenne",
                        site=="SS" ~ "Sheep Station",
                        site=="WI" ~ "Wildcat"),
         environment_text=paste(site, treatment, sep=", ")) %>%
  rbind(
    tibble(
      site = "Growth Chamber",
      density = "Growth Chamber",
      albedo = "Growth Chamber",
      environment = 17,
      gravel = "Growth Chamber",
      treatment = "Growth Chamber",
      environment_text = "Growth Chamber"
    )
  )

selection_tibble <- 
tibble(
  method="KO",
  Variable=Variable,
  environment=environment,
  coef=unlist(map(S_KO, ~apply(.x, 1, mean, na.rm=TRUE))),
  upr=unlist(map(S_KO, ~apply(.x, 1, quantile, 0.95, na.rm=TRUE))),
  lwr=unlist(map(S_KO, ~apply(.x, 1, quantile, 0.05, na.rm=TRUE)))
) %>%
  left_join(Environments_classification, by="environment") %>%
  mutate(environment_text = factor(environment_text, levels = c(
  "Sheep Station, High White", "Sheep Station, Low White", "Sheep Station, High Black", "Sheep Station, Low Black",
  "Cheyenne, High White", "Cheyenne, Low White", "Cheyenne, High Black", "Cheyenne, Low Black",
  "Baltzor, High White", "Baltzor, Low White", "Baltzor, High Black", "Baltzor, Low Black",
  "Wildcat, High White", "Wildcat, Low White", "Wildcat, High Black", "Wildcat, Low Black",
  "Growth Chamber"
))) %>%
  mutate(
    Variable=factor(Variable, levels=unique(c(common_gardens_traits, traits_physiology_study, traits_growth_chamber)))
  )

# Plot with colored facet labels based principal component value 
colors <- c("darkblue", "white", "darkred") # Match colors in other BromeCast manuscripts 
pc1_values = c(-3.2051, -2.0867, 0.3363, 2.3312)
pc1_values_scaled = (pc1_values - min(pc1_values))/sum((pc1_values - min(pc1_values)))+0.25 ### To do scale and center to c(0,1)
env_colors <- scales::gradient_n_pal(colors)
env_colors_labels <- env_colors(pc1_values_scaled)

# Plot with colored facet labels using strip_themed
selection_coefficient_plot <-
  selection_tibble %>%
  ggplot(aes(x = Variable, y = coef, group = interaction(Variable, method))) +
  geom_point(shape = 19, size = 2, color = "black", position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lwr, ymax = upr, color = (lwr > 0 | upr < 0)), 
                width = 0.2, position = position_dodge(width = 0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap2(
    ~environment_text, 
    ncol = 4,
    nrow = 5,
    scales = "free", 
    strip = strip_themed(
      background_x = elem_list_rect(fill = c(rep(env_colors_labels, each=4), "lightgrey"))
    )
  ) +
  theme_bw() +
  labs(x = "", y = "Selection Coefficient", title = element_blank()) +
  scale_color_manual(values = c("black", "red"), guide = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

### Save plot 
ggsave(plot=selection_coefficient_plot, 
       filename=file.path(dir, "Figures", "selection_coefficient.png"),  
       dpi=250, height=10, width=8)
```

## 10-fold cross-validation 

### Create Folds 

```{r, Create Folds 10-fold cross-validation}

###
### Data
###

### Filter for observed traits
data_obs <-
  data_all %>%
  filter(!is.na(value)) %>%
  ### Add experiments for stratification 
  mutate(
    experiment = case_when(
        t %in% 1:6 & e <= 16  ~ "common garden",
        t %in% 7:10 & e %in% c(8, 13) ~ "physiology study",
        t %in% c(1:6, 11:14) & e > 16 ~ "growth chamber"
      )
  )

###
### Create Folds 
###

# Set seed for reproducibility 
set.seed(3134)

# Create the 10-fold cross-validation experiment stratifying by experiment 
folds <- vfold_cv(
  data_obs, 
  v = 10,
  strata = experiment
)

data_folds <- folds$splits %>%
  # Use map to iterate over each rsplit object
  map(function(split) {
    
    # Extract Training Data and label it
    train_data <- training(split) %>%
      mutate(set_type = "train")
    
    # Extract Testing Data and label it
    test_data <- testing(split) %>%
      mutate(set_type = "test")
    
    # Combine them vertically
    bind_rows(train_data, test_data)
  })

# Save 
save(data_folds, file=file.path(dir, "Outputs", "data_folds.RData"))
```

### Run Cross validation

```{r, Run 10-fold cross-validation}


###
### Parallel Cross Validation Study
###

n_folds <- 10

### Things we don't want exported 
packages_list <- c("tidyverse", "Matrix", "abind", "cmdstanr")

num_cores <- 10
cl <- makeCluster(num_cores)
registerDoParallel(cl)
Start_all=Sys.time()
metrics <-
  foreach(m=1:n_folds,
          .packages=packages_list,
          .errorhandling="remove",
          .combine="rbind") %dopar% {
  
            ### Set up storage tibble
            metrics_out <-
              tibble(
                fold=NULL,
                model=NULL,
                data=NULL,
                genotype=NULL,
                trait=NULL,
                environment=NULL,
                time=NULL,
                coverage=NULL,
                bias=NULL,
                sd=NULL,
                ESS=NULL
              )
          
            ### Hyperparameters 
            chains <- 1
            iter_warmup <- 200
            iter_sampling <- 1000
            refresh <- 100
            
            ### Functions 
            source(file.path(dir, "algorithms", "functions.R"))
            
            ### Folds
            load(file=file.path(dir, "Outputs", "data_folds.RData"))
            ### Other Data 
            load(file=file.path(dir, "Outputs", "data_list.RData"))
            data_all <- data_folds[[m]]
            
            ### Extract true values 
            test <- 
              data_all %>%
              filter(set_type == "test") 
            
            ### Now set testing to NA
            data_all <-
              data_all %>%
              mutate(
                value = if_else(set_type == "test", NA, value)
              )

            ### Format Data 
            NAs <- 
              data_all$value %>%
              is.na() %>%
              which()
            
            ### Build the list
            list_all <- list(
              # Dimensions 
              N = length(data_all$value[-NAs]),
              N_p = length(data_all$value[NAs]),
              P = ncol(data_list$X),
              G = max(data_all$g),
              E = max(data_all$e),
              T = max(data_all$t),
              T_a = 6,
              q = 4,
              # In Sample 
              g_idx = data_all$g[-NAs], 
              e_idx = data_all$e[-NAs],   
              t_idx = data_all$t[-NAs],   
              y = data_all$value[-NAs],
              X = data_list$X,
              K = data_list$K[[2]],
              # Out of sample
              g_idx_p = data_all$g[NAs], 
              e_idx_p = data_all$e[NAs],   
              t_idx_p = data_all$t[NAs]
            )
            
            # Define the models to fit (only doing low-rank methods MMM too time consuming)
            models <- c("KO", "CO", "UO", "CCO")

            ###
            ### Fit Ordination Models
            ###
            
            metrics <- map_dfr(models, function(m) {
              
              cat(paste("\n--- Fitting Model:", m, "---\n"))
  
              # Fit the current model (m)
              fit <- fit_model(
                list_all,
                chains = chains,
                dir = dir, 
                iter_warmup = iter_warmup,
                iter_sampling = iter_sampling,
                refresh = refresh,
                model = c(m),  
                pathfinder = TRUE
              )
              
              # Extract performance metrics
              metrics <- extract_performance_CV(
                fit,
                m,
                test
              )
              
              # Return the combined metrics tibble for this model
              return(metrics)
            }) 
            
  

            cat(m, "")
            return(metrics)

}
End_all <- Sys.time() # 3.43 hours 
difftime(End_all, Start_all)

# save metrics 
save(metrics, file=file.path(dir, "Outputs", "metrics_CV10.RData"))

metrics_summary <- metrics %>%
  mutate(
    ESR = log10( ess_bulk / time )
  ) %>%
  group_by(trait, model) %>%
  summarise(
    mean_coverage = mean(as.numeric(coverage)),  # coverage is logical, convert to numeric
    mean_bias     = mean(bias, na.rm = TRUE),
    mean_sd       = mean(sd, na.rm = TRUE),
    mean_ESR      = mean(ESR, na.rm = TRUE),
    .groups = "drop"
  )
```

### Make Cross Validation Plots 

```{r, Make Cross Validation Plots}

###
### Make Plots 
###


### Inputs   
load(file=file.path(dir, "Outputs", "metrics_CV10.RData"))

### Cross validation (average over folds and environments) means
CV_boxblot <-
  metrics %>%
  mutate(esr = log10( ess_bulk / time ),
         trait=factor(trait, levels=c("FT", "Seed(n)", "Seed(m)", "Biomass", "Tiller(n)", "Survival",
                                      "Phenological", "Freeze", "Fv/Fm", "SLA",
                                      "Height", "Growth", "Leaves(n)", "Inflorescence(h)"))) %>% 
  group_by(model, trait, genotype) %>%
  summarise(across(c("sd", "bias", "coverage", "esr"), mean)) %>%
    ### Remove some outliers for cleaner plotting 
  filter(coverage>0.8,
         bias<2.5,
         esr<2) %>%
  gather(key="metric", value="value", -c("model":"genotype")) %>%
  mutate(metric=case_when(metric=="bias" ~ "Bias",
                          metric=="coverage" ~ "Coverage",
                          metric=="sd" ~ "Variability",
                          metric=="esr" ~ "Efficiency"),
         metric=factor(metric, 
                       levels=c("Coverage", "Bias", "Variability", "Efficiency"),
                       labels=c("Coverage", "Bias", "Variability", "Efficiency")),
         model=factor(model, 
                      levels = c("KO", "CCO", "CO", "UO"),
                      labels = c("Kinship Ordination", 
                                 "Concurrent Ordination", 
                                 "Constrained Ordination",
                                 "Unconstrained Ordination"))) %>%
  filter(!is.na(value),
         !is.na(metric)) %>%
  ggplot(aes(x=trait, y=value, fill=model)) +
    facet_wrap(~metric, scales = "free_y", nrow=4) +
    #facet_grid(metric~Sigma, scale = "free") +  # results facet by correlation 
    geom_boxplot(outlier.shape = NA) +
    theme_bw() +
    theme(legend.position = "bottom") +
    xlab("") +
    ylab("") +
    ggtitle("") +
    scale_fill_manual(name="Model", 
                      values=wes_palette("Darjeeling1", 5, type = c("discrete"))[2:5])

ggsave(plot=CV_boxblot, 
       filename=file.path(dir, "Figures", "CV_boxblot.png"),  
       dpi=1000, height=8, width=13)

```

